\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-10 lang-design.tex
took a pass on §3

The \lsti{FInduction} command is similar to \lsti{FRecursion} but
allows cases to be defined in proof mode.
Consider "weakenlem" as an example.
It is proven by induction on the typing relation "hasty".
Its \lsti{motive} clause suggests that the lemma reads as

\begin{centered}
\begin{minipage}{.63\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T.
\end{lstlisting}
\end{minipage}
\end{centered}

@@

\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
  \forall ...
\end{lstlisting}

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by induction on "hasty" and because "STLCFix" adds
a constructor to "hasty", the programmer is required in family "STLCFix" to
extend the proof of "substlem" to handle this extra case.

@@

As another example, consider the type of "st_beta" in family "STLC".
It refers to "subst", whose meaning is late bound.
When "st_beta" is inherited into family "STLCFix", "st_beta" has a type that now
refers to the "subst" function in "STLCFix", where "subst" is defined
by handling all the five cases known to that family.
Thus, late binding of "subst" allows the derived family to reuse "st_beta"
to $\beta$-reduce abstractions whose body is a fixpoint.

Importantly, a name is late bound only within a family that defines or extends it.
Outside such families, the name can be accessed only by explicitly
specifying a family that contains it.
For example, the last line of \cref{fig:stlc-mechanized} accesses "typesafe"
with "STLCFix.typesafe".
Executing the command prints

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLCFix.typesafe: \forall t t' T, STLCFix.steps t t' -> STLCFix.hasty STLCFix.empty t T ->
                   STLCFix.value t' \/ \exists t'', STLCFix.step t' t''.
\end{lstlisting}

@@

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall G' G x t' T', G' = extend G x T' -> hasty empty t' T' -> hasty G (subst tm_unit x t') ty_unit.
\end{lstlisting}

@@

However, with \lsti{subst} being late bound, \lsti{subst} can no longer
be unfolded~(\ref{chg:latebind-v-defeq}):
the definition of \lsti{subst}, as a recursive function,
varies across families, yet a derived family should be able to reuse the
proof of the \lsti{ht_unit} case
even when it has to modify the definition of \lsti{subst}.
Without the ability to unfold \lsti{subst},
how can the programmer make progress in this proof, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{\forall x t, subst tm_unit x t = tm_unit},
as a proposition (\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.\EDJ{I think the reviewer might ask : what if the subst late-bound tm_unit case/why the subst cases cannot be late-bound? it not necessary return tm_unit in those case. 
\\ What the plugin generates is actually subst tm_unit x t = subst.tm_unit. In other words, it points back to the handler instead of the concrete constructor. That is one of the advantage of looking FRecursion as a family.}
%
Based on this insight, \Lang automatically generates
propositional equalities for each recursive function defined with \lsti{FRecursion},
making them available as axioms for use by the rest of the current family.
\Lang also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the postulated
propositional equalities.

Note that the definitional equality on \lsti{subst} is available
outside those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound, so only
propositional equality is available.
In contrast, outside those families, \lsti{subst} is always referenced
by specifying a family that contains it, \EDJ{I think the reviewer without background of Family Polymorphism will be confused at this paragraph. i.e. Why inside family it is late-bound but not outside? }
such as \lsti{STLC.subst} or \lsti{STLCFix.subst}.
As far as the type checker is concerned, \lsti{tm_unit} and
\lsti{STLCFix.subst tm_unit x t'} are the same thing---the type checker
equates them definitionally by unfolding \lsti{STLCFix.subst} and
performing normalization.

@@

Note that the requirement still allows a derived family to introduce
new declarations into the context of a field. %\EDJ{Maybe a better wording is ``the relative order between fields are preserved during inheritance''?}
For example, the left column of \cref{fig:stlc-isorec-prod} shows an
extension of STLC with iso-recursive types, where "tysubst" is
introduced into the context of "hasty".
%
In the event that it cannot be inferred where to place of a newly introduced field,
programmer annotation is required.

@@

In \cref{fig:stlc-isorec-prod}, family "STLCFixIsorec" is declared as
an "STLC" extension that mixes in "STLCFix" and "STLCIsorec".
The family is declared with minimal verbiage, yet
\lsti{STLCFixIsorec.typesafe} is automatically a proof of the type-safety
theorem of an STLC equipped with fixpoints and iso-recursive types.

@@

Partial recursors appear weaker than ordinary recursors,
but there is power in restraint.
Partial recursors offer a principled, uniform way to derive injectivity and disjointness of
constructors, while allowing for extensibility:
they enable injective mappings from a late bound inductive
type, like "tm", to an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first applies the injective mapping to
both sides of~\lsti{H}, obtaining

>>>>

The \lsti{FInduction} command is similar to \lsti{FRecursion} but
allows cases to be defined in proof mode.
Consider "weakenlem" as an example.
It is proven by induction on the typing relation "hasty".
Its \lsti{motive} clause shows that the lemma reads as
\lstinline[basicstyle=\fontsize{8}{9}\ttfamily]!\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T!.

%\begin{centered}
%\begin{minipage}{.63\textwidth}
%\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
%\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T.
%\end{lstlisting}
%\end{minipage}
%\end{centered}

@@

\begin{centered}
\begin{minipage}{.635\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall G', includedin G G' -> hasty G' (tm_abs x t) (ty_arrow T1 T2)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
and with the induction hypothesis
\lstinline[basicstyle=\fontsize{8}{9}\ttfamily]!\forall G', includedin (extend G x T1) G' -> hasty G' t T2!.
%
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by \lsti!FInduction! on "hasty" and because "STLCFix" adds
a constructor "ht_fix" to "hasty", the programmer is required in family "STLCFix" to
extend the proof of "substlem" to handle this extra case.

@@

As another example, consider the type of "st_beta" in family "STLC".
It refers to "subst", whose meaning is late bound.
When "st_beta" is inherited into family "STLCFix", "st_beta" has a type that now
refers to the "subst" function in "STLCFix", where "subst" is defined
by handling all the five cases known to that family.
Thus, late binding of "subst" allows the derived family to reuse "st_beta"
as the $\beta$-reduction rule for applications possibly constructed from "tm_fix".

Importantly, a name is late bound only within a family that defines or further binds it.
Outside such families, the name can be accessed only by explicitly
specifying a family that contains it.
For example, the last line of \cref{fig:stlc-mechanized} accesses "typesafe"
with a qualifier "STLCFix".
The command prints

\begin{centered}
\begin{minipage}{.880\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLCFix.typesafe: \forall t t' T, STLCFix.steps t t' -> STLCFix.hasty STLCFix.empty t T ->
                  STLCFix.value t' \/ \exists t'', STLCFix.step t' t''.
\end{lstlisting}
\end{minipage}
\end{centered}

@@

\begin{centered}
\begin{minipage}{\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall G' G x t' T', G' = extend G x T' -> hasty empty t' T' -> hasty G (subst tm_unit x t') ty_unit.
\end{lstlisting}
\end{minipage}
\end{centered}

@@

But with \lsti{subst} being late bound, \lsti{subst} cannot and should not
be unfolded~(\ref{chg:latebind-v-defeq}):
the definition of \lsti{subst}, as a recursive function,
varies across families, yet a derived family should be able to reuse the
proof of the \lsti{ht_unit} case
even when it has to modify the definition of \lsti{subst}.
Without the ability to unfold \lsti{subst},
how can the programmer make progress in this proof, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{\forall x t, subst tm_unit x t = tm_unit},
as a proposition (\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%\EDJ{I think the reviewer might ask : what if the subst late-bound tm_unit case/why the subst cases cannot be late-bound? it not necessary return tm_unit in those case. 
%\\ What the plugin generates is actually subst tm_unit x t = subst.tm_unit. In other words, it points back to the handler instead of the concrete constructor. That is one of the advantage of looking FRecursion as a family.}\YZreply{Currently the paper is taking the position that case handlers cannot be further bound, to make the story simpler.}
What varies is \lsti{subst} itself as a recursive function combining the case handlers.
%
Based on this insight, \Lang automatically generates
a propositional equality for each case handler defined within \lsti{FRecursion},
making the equalities and the recursive function available as axioms for use by the rest of the current family.
\Lang also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting applications of the
axiomatized recursive function using the axiomatized propositional equalities.

Note that the definitional equality on \lsti{subst} is available
\emph{outside} those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound
(i.e., polymorphic to the enclosing family), so only
propositional equality is available.
In contrast, outside those families, \lsti{subst} is always referenced
by specifying a family that contains it, %\EDJ{I think the reviewer without background of Family Polymorphism will be confused at this paragraph. i.e. Why inside family it is late-bound but not outside? }
as in \lsti{STLC.subst} and \lsti{STLCFix.subst}.
As far as the type checker is concerned, \lsti{tm_unit} and
\lsti{STLCFix.subst tm_unit x t'} are the same thing---the type checker
equates them definitionally by unfolding \lsti{STLCFix.subst} and
performing normalization.

@@

Note that the requirement still allows a derived family to introduce
new declarations into the context of a field. %\EDJ{Maybe a better wording is ``the relative order between fields are preserved during inheritance''?}
For example, the left column of \cref{fig:stlc-isorec-prod} shows an
extension of STLC with iso-recursive types, where "tysubst" is
introduced into the context of "hasty".
%
In the event that \Lang cannot infer where the programmer intends to place a new field,
annotation is required.

@@

In \cref{fig:stlc-isorec-prod}, "STLCFixIsorec" is declared as
an "STLC" extension that mixes in "STLCFix" and "STLCIsorec".
The family is declared with minimal verbiage, yet
\lsti{STLCFixIsorec.typesafe} is automatically a proof of the type-safety
theorem of an STLC equipped with fixpoints and iso-recursive types.

@@

Partial recursors appear weaker than ordinary recursors,
but there is power in restraint.
Partial recursors serve as extensional characterizations of inductive types allowing future extension.
In particular,
they offer a principled, uniform way to derive injectivity and disjointness of
constructors:
they enable injective mappings from a late bound inductive
type, like "tm", to an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first applies an injective mapping to
both sides of~\lsti{H}, obtaining
